---
title: 'TICTAC: Target illumination clinical trials analytics with cheminformatics'
author: "Jeremy Yang"
output:
  html_document:
    number_sections: yes
    toc: yes
editor_options: 
  chunk_output_type: console
---

# Introduction

Mining ClinicalTrials.gov for Target Hypotheses: AACT Drug-Targets Analysis

ClinicalTrials.gov was first released in 2000. As of March 2019, ClinicalTrials.gov includes 300,676 research studies in all 50 states and in 208 countries.   The the CTTI AACT project and database provides a harmonizing schema and convenient access. However, there remain major challenges to knowledge discovery using these data, such as lack of standard terminology.  To address this, for the use case of elucidating drug target hypotheses, we have used state of the art domain specialized text mining with synonym resolution for specific classes of entities: (1) chemicals and (2) diseases. Chemicals are identified and resolved using NextMove Leadmine. Diseases, indications and other phenotypic terms are mined via JensenLab Tagger with Disease Ontology dictionary, and NLM supplied MeSH terms. Protein targets are associated via ChEMBL bioactivities on molecular structure cross-referencing. Another fundamental challenge is to assess confidence of inferences from noisy and disparate data. We propose a scoring system for assessing confidence for target hypotheses inferred from aggregated clinical trials, with emphasis on higher confidence, novel predictions with the potential to illuminate the understudied druggable genome. 

* CTTI = Clinical Trials Transformation Initiative
* AACT = Aggregate Analysis of ClinicalTrials.gov
* According to website (2022), data is refreshed daily.
* <https://aact.ctti-clinicaltrials.org/>
* <https://github.com/ctti-clinicaltrials/aact>
* <https:/www.nextmovesoftware.com>
* <https://bitbucket.org/larsjuhljensen/tagger>
* <https://download.jensenlab.org/>

## Issues

* Prior belief is that target NER is not likely to be useful, since clinical trials descriptive text is not 
generally written to communicate molecular mechanisms to research scientists, but with focus on _clinical_ 
efficacy and safety. In due diligence we perform target NER, and to quantify concordance or refutation with
our prior belief, we also perform target NER on arbitrary non-biomedical text, tweets from the Twitter API for
__\#brexit__ (26 Nov 2019). We find that 8.64 target entities per 1000 chars in the tweets, vs. 6.63 in the clinical trials
descriptions. While not proof this does support our belief and less direct method via chemical NER. 

## To do

* We intend to calculate multivariate &mu; scores to assess and rank disease-target
associations. So identification of suitable evidence variables are needed.

## Identifier mappings:

> `NCT_ID` &rarr;(JensenLab:Tagger)&rarr; `DOID`  
> `NCT_ID` &rarr;(AACT)&rarr; `MeSH`  
> `NCT_ID` &rarr;(NextMove:LeadMine)&rarr; `SMILES`  
> `SMILES` &rarr;(PubChem)&rarr; `CID`  
> `CID` &rarr;(PubChem)&rarr; `INCHIKEY`  
> `INCHIKEY` &rarr;(ChEMBL)&rarr; `MOLECULE_CHEMBL_ID`  
> `MOLECULE_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `ACTIVITY_ID`  
> `ACTIVITY_ID` &rarr;(ChEMBL)&rarr; `TARGET_CHEMBL_ID`  
> `TARGET_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `COMPONENT_ID`  
> `COMPONENT_ID` &rarr;(ChEMBL)&rarr; `UNIPROT`  
> `ACTIVITY_ID` &rarr;(ChEMBL)&rarr; `DOCUMENT_CHEMBL_ID`  
> `DOCUMENT_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `PUBMED_ID`  

## Input files:

* (AACT) `aact_studies.tsv`
* (AACT) `aact_drugs.tsv`
* (AACT) `aact_descriptions.tsv`
* (NextMove LeadMine) `aact_drugs_leadmine.tsv`
* (PubChem) `aact_drugs_smi_pubchem_cid.tsv`
* (PubChem) `aact_drugs_smi_pubchem_cid2ink.tsv`
* (ChEMBL) `aact_drugs_ink2chembl.tsv`
* (ChEMBL) `aact_drugs_chembl_activity.tsv`
* (ChEMBL) `aact_drugs_chembl_target_component.tsv`
* (ChEMBL) `aact_drugs_chembl_document.tsv`
* (IDG TCRD/Pharos) `pharos_targets.tsv`
* (JensenLab Tagger) `aact_descriptions_tagger_disease_matches.tsv`
* (JensenLab Dictionary) `diseases_entities.tsv`

> `nct_id` is the study ID.

```{r echo=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
base::date()
```

```{r setup, echo=FALSE, message=FALSE}
library(readr)
library(data.table)
library(plotly, quietly=T)
```

```{r}
truncate.ellipsis <- function(txt, maxlen) {
  return (ifelse(nchar(txt)<maxlen, txt, sprintf("%s...", substr(txt, 1, maxlen-3))));
}
```
# AACT Db Timestamp
```{r}
dbts <- as.Date(read.table("data/aact_timestamp.txt")[[1]], "%Y-%m-%d")
sprintf("AACT database timestamp: %s", dbts)
```

# Input studies and drugs
## Studies
Read file of all studies in AACT.

```{r studies_input, message=FALSE}
studies <- read_delim("data/aact_studies.tsv", "\t")
setDT(studies)
studies[["start_year"]] <- as.integer(format(studies$start_date, "%Y"))
n_studies_total <- nrow(studies)
sprintf("Total studies: %d ; unique NCT_IDs: %d", n_studies_total, uniqueN(studies$nct_id))
```

### Study references
Reference type `results_reference` may offer greater evidence, confidence.

```{r}
refs <- read_delim("data/aact_study_refs.tsv", "\t", col_types="ccccc")
setDT(refs)
refs[, id:=NULL]
sprintf("references: %d; NCT_IDs: %d; PMIDs: %d; results_references: %d", nrow(refs), 
        uniqueN(refs$nct_id),
        uniqueN(refs$pmid),
        nrow(refs[reference_type=="results_reference"]))
refs <- refs[, .(N_refs = .N, N_results_refs = sum(reference_type=="results_reference"), pmids=paste0(pmid, collapse="; "), citations=paste0(citation, collapse="\t")), by="nct_id"]
```

## Drugs 
Read file of all drugs in AACT.

* `id` is AACT `INTERVENTION_ID`, corresponding with an instance of a drug, dose, delivery, etc. in a study.
* Note that one study may involve multiple drugs.
* At this point a "drug" is imprecisely identified by name, generally one of many synonyms.

```{r message=FALSE}
drugs <- read_delim("data/aact_drugs.tsv", "\t", col_types = "dcc")
setDT(drugs)
drugs <- drugs[order(drugs$name)]
sprintf("Unique drug names: %d ; unique intervention IDs: %d", uniqueN(drugs$name), uniqueN(drugs$id))
```

## Studies: Interventional drug studies only
Select only ___Interventional___ studies (`study_type`) associated with drugs (via `NCT_ID`). No __Observational__ this analysis.

```{r}
studies <- studies[study_type=="Interventional"]
n_studies_inter <- nrow(studies)
sprintf("Interventional studies: %d (%.1f%%)", n_studies_inter, 100*n_studies_inter/n_studies_total)
studies$phase[studies$phase == "N/A"] <- NA
studies <- studies[nct_id %in% drugs$nct_id]
n_studies_inter_drug <- nrow(studies)
sprintf("Interventional drug studies: %d ; unique NCT_IDs: %d", n_studies_inter_drug, uniqueN(studies$nct_id))
```


```{r}
drugs <- merge(drugs, studies, by="nct_id", all.x=T, all.y=F)
knitr::kable(merge(studies[, .(N_studies = .N), by="phase"], drugs[, .(N_drugs = .N), by="phase"], by="phase")[order(phase)], caption="Drug studies and drugs, by phase")
```

```{r}
knitr::kable(merge(studies[, .(N_studies = .N), by="overall_status"], drugs[, .(N_drugs = .N), by="overall_status"], by="overall_status")[order(-N_studies)], caption="Drug studies and drugs, by overall_status")
```

## Drug studies by Phase and Status

```{r message=FALSE}
ax0 <- list(showline=F,zeroline=F,showticklabels=F,showgrid=F)
plot_ly() %>%
  add_pie(data=drugs[, .(.N), by="phase"], labels=~phase, values=~N, sort=F,
          textinfo="label+percent", textposition="inside", domain=list(x=c(0, 0.5), y=c(0, 1))) %>%
  add_pie(data = drugs[, .(.N), by="overall_status"], labels=~overall_status, values=~N,
          textinfo = "label+percent", textposition="inside", domain=list(x = c(0.5, 1), y = c(0, 1))) %>%
  add_annotations(x = c(0.25, 0.75), y = c(-.1, -.1),
                  xanchor = "center", xref = "paper", yref = "paper", showarrow = F,
                  text = c("Phase", "Overall status"),
                  font = list(family = "Arial", size = 20)) %>%
  layout(title=paste0("AACT: Drug trials by phase and status<br>(N_total = ", nrow(drugs), ")"),
         xaxis=ax0, yaxis=ax0, margin=list(t=120), font=list(family="Arial", size=20), showlegend=F)
```

## Drug studies and drugs by start_year

```{r message=FALSE}
yi <- 1990
yf <- 2021
plot_ly(type="bar", data=merge(studies[, .(N_studies = .N), by="start_year"], drugs[, .(N_drugs = .N), by="start_year"], by="start_year"), x = ~start_year, name="studies", y = ~N_studies) %>%
	add_trace(name="drugs", y = ~N_drugs) %>%
	layout(title=sprintf("AACT: Studies and drugs %d-%d", yi, yf), xaxis=list(range=c(yi, yf), tickangle=-45), yaxis=list(title="N"),
	       barmode="stack",
	       margin=list(t=120), font=list(family="Monospace", size=20),
	       legend=list(x=0.1, y=1))
```

# NextMove Leadmine Chemical NER
AACT drug names resolved to standard names and structures via SMILES.
Note that one name may include multiple chemicals.
Now we can use cheminformatically rigorous counts for drugs as active pharmaceutical ingredients (APIs).

```{r message=FALSE}
cner <- read_delim("data/aact_drugs_leadmine.tsv",  "\t", escape_double=F, trim_ws=T)
setDT(cner)
setnames(cner, old=c("DocName", "ResolvedForm"), new=c("id","smiles"))
sprintf("Drug unique SMILES resolved by LeadMine: %d ; unique intervention IDs: %d; unique names: %d", uniqueN(cner$smiles), uniqueN(cner$id), uniqueN(cner$OriginalText))
```
## Chemical NER mentions
### Totals by merging of synonyms to resolved structure (locally canonical SMILES)

```{r}
cner_totals <- cner[!is.na(smiles), .(N_mentions = .N), by=c("smiles", "OriginalText")]
cner_totals <- cner_totals[order(smiles, OriginalText)]
cner_totals <- cner_totals[, .(N_mentions = sum(N_mentions), names=paste0(OriginalText, collapse="; ")), by=c("smiles")]
cner_totals <- cner_totals[order(-N_mentions)]
cner_totals[["smi2img"]] <- sprintf("<img src=\"https://datascience.unm.edu/tomcat/cdkdepict/depict/cow/svg?smi=%s&h=50&w=80&zoom=3\">", sapply(cner_totals$smiles, URLencode, reserved=T))
write_delim(cner_totals, "data/cner_totals.tsv", "\t")
knitr::kable(cner_totals[1:20, .(smi2img, N_mentions, names)], caption="Top 20 drugs by total mentions")
```

### Chemical NER mentions resolved to structures (SMILES)

```{r}
drugs <- merge(drugs, cner, by="id")
drugs[["resolved_structure"]] <- !is.na(drugs$smiles)
message(sprintf("Drugs (drug names) with resolved structure: %d / %d (%.1f%%)", sum(drugs$resolved_structure), nrow(drugs), 100*sum(drugs$resolved_structure)/nrow(drugs)))
```

### Chemical NER mentions by intervention ID.

```{r}
cner_this <- cner[!is.na(smiles), .(.N), by="id"]
sprintf("Mentions by intervention ID: %d / %d (%.1f%%)", nrow(cner_this), uniqueN(drugs$id), 100*nrow(cner_this)/uniqueN(drugs$id))
```

### Chemical NER mentions by trial (NCT ID).

```{r}
cner <- merge(cner, drugs[, .(id, nct_id)], by="id")
cner_this <- cner[!is.na(smiles), .(.N), by="nct_id"]
sprintf("Mentions by study: %d / %d (%.1f%%)", nrow(cner_this), uniqueN(drugs$nct_id), 100*nrow(cner_this)/uniqueN(drugs$nct_id))
```

### Chemical NER mentions by drug, i.e. name in AACT.

```{r}
cner_this <- cner[!is.na(smiles), .(.N), by="OriginalText"]
sprintf("Mentions by drug name: %d / %d (%.1f%%)", nrow(cner_this), uniqueN(drugs$name), 100*nrow(cner_this)/uniqueN(drugs$name))
```

# PubChem:

## Intervention IDs to CIDs from PubChem

SMILES mapping normally problematic without canonicalization, but we use the same SMILES strings, generated by NextMove,
and associated with CIDs via PubChem REST API.

```{r message=FALSE}
cid2ink <- read_delim("data/aact_drugs_smi_pubchem_cid2ink.tsv", "\t")
setDT(cid2ink)
sprintf("PubChem CIDs with InChIKeys: %d", cid2ink[, uniqueN(CID)])
smi2cid <- read_delim("data/aact_drugs_smi_pubchem_cid.tsv", "\t", col_types = cols(CID=col_integer()))
setDT(smi2cid)
sprintf("PubChem SMILES2CID hits: %d / %d (%.1f%%)", sum(!is.na(smi2cid$CID)), nrow(smi2cid), 100*sum(!is.na(smi2cid$CID))/nrow(smi2cid))
```

```{r}
drug2cid <- unique(merge(smi2cid, unique(drugs[, .(smiles, id)]), all.x=F, all.y=F, by.x="SMILES", by.y="smiles"))
setnames(drug2cid, old="id", new="itv_id")
drug2cid <- drug2cid[!is.na(CID) & CID!=0]
drug2cid <- drug2cid[, `:=`(SMILES=NULL, Name=NULL)]
sprintf("Intervention IDs mapped to PubChem CIDs (via SMILES): %d", drug2cid[, uniqueN(itv_id)])
write_delim(drug2cid, "data/aact_drugs_itvid2cid.tsv", delim="\t")
```

# IDG/TCRD:
For Target Development Level (TDL) and other metadata.

```{r message=FALSE}
tcrd_tgt <- read_delim("data/tcrd_targets.tsv", "\t", col_types=cols(.default=col_character()))
setDT(tcrd_tgt)
setnames(tcrd_tgt, old=c("TDL", "tcrdTargetFamily"), new=c("idgTDL", "idgFamily"))
```

# ChEMBL:
## ChEMBL molecule IDs, and properties (via InChIKeys)
Perhaps should instead use PubChem CIDs and UniChem.

```{r}
chembl_mol <- read_delim("data/aact_drugs_ink2chembl.tsv", "\t", 
                         col_types=cols_only(molecule_chembl_id = col_character(), molecule_type = col_character(), structure_type = col_character(), pref_name=col_character(), max_phase=col_character(), therapeutic_flag=col_logical()))
sprintf("ChEMBL compounds mapped via InChIKeys: %d", nrow(chembl_mol))
```

## ChEMBL activities for mapped compounds
Select only activities with pChembl values for relevance to protein targets and confidence.

```{r message=FALSE}
chembl_act <- read_delim("data/aact_drugs_chembl_activity.tsv", "\t", col_types=cols(
  .default = col_character(),
  activity_id = col_double(),
  document_year = col_integer(),
  pchembl_value = col_double(),
  standard_value = col_double(),
  target_tax_id = col_integer(),
  text_value = col_logical(),
  value = col_double()))
setDT(chembl_act)
sprintf("ChEMBL activities: %d", nrow(chembl_act))
sprintf("ChEMBL activities molecules: %d ; canonical_smiles: %d ; targets: %d ; documents: %d", uniqueN(chembl_act$molecule_chembl_id), uniqueN(chembl_act$canonical_smiles), uniqueN(chembl_act$target_chembl_id), uniqueN(chembl_act$document_chembl_id))
plot_ly(x=chembl_act$pchembl_value, type="histogram", nbinsx=20) %>%
  layout(title=paste0("ChEMBL Activities, pChembl values<br>(N_total = ", nrow(chembl_act), ")"), 
         xaxis=list(title="pChembl"), yaxis=list(title="N"),
         margin=list(t=140), font=list(family="Monospace", size=20),
         annotations=list(text="pChembl = -log<sub>10</sub>({I|E|L}C50})", showarrow=F, x=1, y=0.9, xref="paper", yref="paper", xanchor="right", font=list(size=14)),
         showlegend=F)
```

### Activity and molecule counts by assay types

```{r}
atypes <- c(B = "Binding", F = "Functional", A = "ADMET", T = "Toxicity", P = "Physicochemical", U = "Unclassified")
atype_counts <- chembl_act[, .(N_molecule = uniqueN(molecule_chembl_id), N_activity = .N), by="assay_type"][order(-N_molecule)]
atype_counts[, assay_type := sprintf("%s:%s", assay_type, atypes[assay_type])]
knitr::kable(atype_counts, caption="Activity and molecule counts by assay types")
```

### Activity and molecule counts by assay source
```{r}
chembl_srcs <- read_delim("data/chembl_sources.tsv", "\t", col_types=cols(.default=col_character()))
setDT(chembl_srcs)
asrc_counts <- chembl_act[, .(N_molecule = uniqueN(molecule_chembl_id), N_activity = .N), by="src_id"]
asrc_counts <- merge(chembl_srcs[, .(src_id, src_short_name)], asrc_counts, by="src_id")
asrc_counts <- asrc_counts[order(-N_molecule)]
knitr::kable(asrc_counts[, .(Source = src_short_name, N_molecule, N_activity)], caption="Activity and molecule counts by assay sources")
```

## ChEMBL targets (via activities)

```{r message=FALSE}
chembl_tgt <- read_delim("data/aact_drugs_chembl_target_component.tsv", "\t")
setDT(chembl_tgt)
setnames(chembl_tgt, old="accession", new="uniprot")
sprintf("ChEMBL target proteins: %d", nrow(chembl_tgt))
tgt <- merge(chembl_tgt, tcrd_tgt, all.x=T, all.y=F, by.x="uniprot", by.y="uniprotId")
sprintf("ChEMBL target proteins mapped to TCRD (human): %d", nrow(tgt[!is.na(tgt$idgTDL),]))
```

## ChEMBL targets by organism:

```{r}
sprintf("Organisms: %d", uniqueN(tgt$organism))
org_counts <- tgt[, .(N_targets=.N, Types=paste0(sort(unique(target_type)), collapse="; ")), by="organism"][order(-N_targets)]
knitr::kable(org_counts[1:10], caption="Targets by organism (top 10)")
```

## ChEMBL human single-protein targets only, by IDG family.

```{r}
tgt <- tgt[organism=="Homo sapiens"]
sprintf("Human targets: %d", nrow(tgt))
tgt <- tgt[target_type=="SINGLE PROTEIN"]
sprintf("Human single-protein targets: %d ; unique UniProts: %d", nrow(tgt), uniqueN(tgt$uniprot))
fam_counts <- tgt[, .(.N), by="idgFamily"]
knitr::kable(fam_counts[order(-N)])
plot_ly(type="bar", data=fam_counts, x=~idgFamily, y=~N, marker=list(color="maroon"), text=~N, textposition="auto") %>%
	layout(title="Targets by Family<br>(Human only)",
	       xaxis=list(title="TDL"), margin=list(t=120), font=list(family="Monospace", size=20))
```

## Targets by IDG TDL:

```{r}
tdls <- c("Tdark", "Tbio", "Tchem", "Tclin")
tdl_counts <- tgt[, .(.N), by="idgTDL"]
tdl_counts$tdl_color <- ifelse(tdl_counts$idgTDL=="Tdark", "gray", ifelse(tdl_counts$idgTDL=="Tbio", "red", ifelse(tdl_counts$idgTDL=="Tchem", "green", ifelse(tdl_counts$idgTDL=="Tclin", "blue", NA))))
tdl_counts$idgTDL <- factor(tdl_counts$idgTDL, levels=tdls)
tdl_counts$tdl_color <- factor(tdl_counts$tdl_color, levels=c("gray", "red", "green", "blue"))
tdl_counts <- tdl_counts[order(idgTDL)]
message(sprintf("%8s: %6d", tdl_counts$idgTDL, tdl_counts$N))
plot_ly(type="bar", x=tdl_counts$idgTDL, y=tdl_counts$N, marker=list(color=tdl_counts$tdl_color), text=tdl_counts$N, textposition="auto") %>%
	layout(title="Targets by IDG TDL<br>(Human only)",
	       xaxis=list(title="TDL"), margin=list(t=120), font=list(family="Monospace", size=20))
```

# JensenLab Tagger Diseases NER
With JensenLab DOID entities dictionary.
On descriptions from detailed_descriptions table.

* `serialno` corresponds with DOID.
* `id` is AACT primary key.
Tagger results TSVs lack headings.

```{r message=FALSE, echo=FALSE}
dner <- read_delim("data/aact_descriptions_tagger_disease_matches.tsv", "\t", 
                   col_names = c("id", "term", "serialnm" ), col_types="c----c-c")
dent <- read_delim(paste0(Sys.getenv("HOME"), "/../data/JensenLab/data/diseases_entities.tsv"), "\t", col_names=c("serialnm", "doid"), col_types="c-c")
setDT(dner)
setDT(dent)
dner <- merge(dner, dent, by="serialnm", all.x=T, all.y=F)
descs <- read_delim("data/aact_descriptions.tsv", "\t", col_types=cols(.default=col_character()), col_names=c("id", "nct_id", "description"), escape_double=F, trim_ws=T)
dner <- merge(dner, descs, by="id", all.x=T, all.y=F)
dner[, id:=NULL]
dner[, serialnm:=NULL]
dner <- dner[grepl("^DOID:", doid)] #Why not? "AmyCo:8" etc.? TBE.
dbad <- c("disease", "can", "face")
dner <- dner[!(tolower(term) %in% dbad)]
sprintf("Total disease mentions: %d (in %d studies)", nrow(dner), uniqueN(studies$nct_id))
```

## Disease mention totals by merging to resolved Disease Ontology term (DOID).

```{r}
dner_totals <- dner[, .(N_mentions = .N), by=c("doid", "term")]
dner_totals <- dner_totals[order(as.integer(sub("DOID:", "", doid)), term)]
dner_totals <- dner_totals[, .(N_mentions = sum(N_mentions), terms=paste0(term, collapse="; ")), by=c("doid")]
dner_totals <- dner_totals[order(-N_mentions)]
dner_totals$terms <- truncate.ellipsis(dner_totals$terms, 200)
knitr::kable(dner_totals[1:20], caption="Top 20 diseases by total mentions")
```

## Disease mentions by study.
Sort synonyms terms by frequency.

```{r}
dner <- dner[, .(N_mentions = .N), by=c("nct_id", "doid", "term")]
dner <- dner[order(nct_id, -N_mentions, -as.integer(sub("DOID:", "", doid)))]
dner <- dner[, .(N_mentions = sum(N_mentions), disease_terms=paste0(term, collapse=";")), by=c("nct_id", "doid")]
knitr::kable(dner[nct_id %in% base::sample(unique(dner$nct_id), 10)], caption="Disease mentions by study (Random sample of studies)")
```

# Enumerate ___study-drug-disease-target___ links.

And include references.

Since each study may be associated with multiple drugs, targets and diseases, we 
build a table of all associated combinations, then aggregate by study (`NCT_ID`).
For DOIDs with multiple terms, keep only most common term for simplicity.

```{r}
sddt_links <- copy(dner)
sddt_links[, `:=`(disease_term = sub(";.*$", "", disease_terms), disease_terms = NULL, N_mentions = NULL)]
#sddt_links <- sddt_links[, .(N_disease_mentions = paste0(N_mentions, collapse="; "), disease_terms=paste0(disease_term, collapse="; "), doids = paste0(doid, collapse="; ")), by=c("nct_id")]
sprintf("study-disease links: %d", nrow(sddt_links))
```

## Link to drugs

> `NCT_ID` &rarr;(NextMove:LeadMine)&rarr; `SMILES`  
> `SMILES` &rarr;(PubChem)&rarr; `CID`  

Keep only studies including both disease and drug mentions.

```{r}
drug_links <- unique(drugs[!is.na(smiles), .(nct_id, itv_id = id, drug_name = EntityText)])
drug_links <- merge(drug_links, drug2cid, by="itv_id", allow.cartesian=T)
#drug_links <- drug_links[order(nct_id, smiles, EntityText, nchar(name))]
#drug_links <- drug_links[, .(N_chem_mentions=length(name), PUBCHEM_CIDs = paste0(cid, collapse="; "), names = paste0(name, collapse="; "), EntityTexts = paste0(EntityText, collapse="; "), smiles = paste0(smiles, collapse="; ")), by="nct_id"]
sddt_links <- merge(sddt_links, drug_links, by="nct_id")
sddt_links <- sddt_links[!is.na(doid) & !is.na(CID)]
sprintf("study-drug-disease links: %d", nrow(sddt_links))
sprintf("studies with drug-disease links: %d", uniqueN(sddt_links$nct_id))
```
## ChEMBL links to targets

> `ACTIVITY_ID` &rarr;(ChEMBL)&rarr; `TARGET_CHEMBL_ID`  
> `TARGET_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `COMPONENT_ID`  
> `COMPONENT_ID` &rarr;(ChEMBL)&rarr; `UNIPROT`  

```{r message=FALSE}
sprintf("ACTIVITY_IDs: %d ; TARGET_CHEMBL_IDs: %d ; pairs: %d",
	uniqueN(chembl_act$activity_id), uniqueN(chembl_act$target_chembl_id), nrow(chembl_act))
tgt2tc <- read_delim("data/aact_drugs_chembl_target_component.tsv", "\t")
setDT(tgt2tc)
setnames(tgt2tc, old="accession", new="uniprot")
sprintf("COMPONENT_IDs: %d ; TARGET_CHEMBL_IDs: %d ; pairs: %d",
	uniqueN(tgt2tc$component_id), uniqueN(tgt2tc$target_chembl_id), nrow(tgt2tc)) 
sprintf("UNIPROTs: %d ; SINGLE_PROTEIN UNIPROTs: %d",
	uniqueN(tgt2tc$uniprot), uniqueN(tgt2tc[target_type=="SINGLE PROTEIN", .(uniprot)]))
```

## PubChem molecules to ChEMBL targets

> `CID` &rarr;(PubChem)&rarr; `INCHIKEY`  
> `INCHIKEY` &rarr;(ChEMBL)&rarr; `MOLECULE_CHEMBL_ID`  
> `MOLECULE_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `ACTIVITY_ID`  

```{r message=FALSE}
cid2ink <- read_delim("data/aact_drugs_smi_pubchem_cid2ink.tsv", "\t")
setDT(cid2ink)
sprintf("CIDs: %d ; INCHIKEYs: %d ; pairs: %d",
	uniqueN(cid2ink$CID), uniqueN(cid2ink$InChIKey), nrow(cid2ink)) 
ink2chembl <- read_delim("data/aact_drugs_ink2chembl.tsv", "\t", col_types=cols(.default=col_character(), first_approval=col_integer(), usan_year=col_integer(), withdrawn_year=col_integer()))
setDT(ink2chembl)
sprintf("INCHIKEYs: %d ; MOLECULE_CHEMBL_IDs: %d ; pairs: %d",
	uniqueN(ink2chembl$inchikey), uniqueN(ink2chembl$molecule_chembl_id), nrow(ink2chembl)) 
sprintf("MOLECULE_CHEMBL_IDs: %d ; TARGET_CHEMBL_IDs: %d ; ACTIVITY_IDs: %d ; DOCUMENT_CHEMBL_IDs: %d",
	uniqueN(chembl_act$molecule_chembl_id), uniqueN(chembl_act$target_chembl_id),
	uniqueN(chembl_act$activity_id), uniqueN(chembl_act$document_chembl_id))
mol2tgt <- merge(cid2ink, ink2chembl, by.x="InChIKey", by.y="standard_inchi_key")
mol2tgt <- unique(mol2tgt[, .(CID, molecule_chembl_id)])
mol2tgt <- unique(merge(mol2tgt, chembl_act[, .(molecule_chembl_id, target_chembl_id)], by="molecule_chembl_id"))
mol2tgt <- unique(merge(mol2tgt, tgt2tc[, .(target_chembl_id, uniprot)], by="target_chembl_id"))
sprintf("CID2UNIPROT links: %d ; CIDs: %d ; UNIPROTs: %d", nrow(mol2tgt), uniqueN(mol2tgt$CID), uniqueN(mol2tgt$uniprot))
```

### Disease-Target links (via study, drug)

### TDL counts

```{r}
sddt_links <- unique(merge(sddt_links, mol2tgt, by="CID", allow.cartesian=T))
sprintf("study-drug-disease-target links: %d", nrow(sddt_links))
sprintf("studies: %d ; drugs: %d ; diseases: %d ; targets: %d",
uniqueN(sddt_links$nct_id), uniqueN(sddt_links$CID), uniqueN(sddt_links$doid),
uniqueN(sddt_links$uniprot))
sddt_links <- merge(sddt_links, tgt[, .(uniprot, gene_symbol = tcrdGeneSymbol, idgTDL)], by="uniprot")
#write_delim(sddt_links, "data/sddt_links.tsv", "\t")
knitr::kable(unique(sddt_links[, .(uniprot, idgTDL)])[, .(.N), by="idgTDL"], caption="TDL counts")
```

Sample Tdark for interesting-ness.
```{r}
sddt_links_sample <- sddt_links[idgTDL=="Tdark", .(nct_id, drug_name, CID, disease_term, doid, gene_symbol, uniprot, idgTDL)]
sddt_links_sample <- sddt_links_sample[sample(1:nrow(sddt_links_sample), 25)]
knitr::kable(sddt_links_sample, caption=sprintf("Sample of
study-drug-disease-target links (N_total = %d)", nrow(sddt_links)))
```

## PubMed references from AACT studies.

```{r message=FALSE}
aact_refs <- read_delim("data/aact_study_refs.tsv", "\t", col_types = cols(.default = col_character()))
sprintf("Study references: %d ; PMIDs: %d ; studies: %d",
	nrow(aact_refs), uniqueN(aact_refs$pmid), uniqueN(aact_refs$nct_id))
```

## PubMed references from ChEMBL activities.

> `ACTIVITY_ID` &rarr;(ChEMBL)&rarr; `DOCUMENT_CHEMBL_ID`  
> `DOCUMENT_CHEMBL_ID` &rarr;(ChEMBL)&rarr; `PUBMED_ID`  

```{r message=FALSE}
docs <- read_delim("data/aact_drugs_chembl_document.tsv", "\t", col_types = cols(.default = col_character(), year = col_integer()))
sprintf("DOCUMENT_CHEMBL_IDs:: %d ; PMIDs: %d",
	uniqueN(docs$document_chembl_id), uniqueN(docs$pubmed_id))
```

# Aggregating, scoring and ranking ___disease, target___ associations.

Evidence variables:

* nStudy : Study count for association.
* nStudyNewness : Study count weighted by newness of study (newer better).
* nStudyPhase : Study count weighted by phase of study (completed better).
* nPub : Study publications.
* nPubTypes : Study publications (results better).
* nDiseaseMention : Disease mention count for disease-target association.
* nDrugMention : Drug mention count for disease-target association.
* nDrug : Drug count for disease-target association.
* nAssay : Assay count for drug-target association.
* nAssayPchembl : Assay count for drug-target association, weighted by pChembl.

---

